Concurrent Programming with CXL

### 2020-06-16 book/paper.tex
Adding a glossary

\newglossaryentry{atomicity}
{
  name=atomicity,
  description={describes that a certain machine instruction or sequence of machine
    instructions by a process is executed indivisibly and cannot be interleaved with 
    machine instructions of another process}
}
\newglossaryentry{concurrent execution}
{
  name=concurrent execution,
  description={(aka parallel execution) is when there are multiple processes executing and
      their machine instructions are interleaved in an unpredictable manner}
}
\newglossaryentry{context}
{
  name=context,
  description={(aka continuation) describes the state of a running process,
    including its program counter, the values of its variables (stored in its
    register), and the contents of its stack}
}
\newglossaryentry{critical section}
{
  name=critical section,
  description={(aka critical region) is a set of instructions that only one process
    is allowed to execute at a time.  The instructions are, however, not executed
    atomically, as other processes can continue to execute and access shared
    variables}
}
\newglossaryentry{determinism}
{
  name=determinism,
  description={is when the outcome of an execution is uniquely determined by the
    initial state}
}
\newglossaryentry{invariant}
{
  name=invariant,
  description={is a binary predicate over states that must hold for every
  reachable state of a process}
}
\newglossaryentry{machine instruction}
{
  name=machine instruction,
  description={is an atomic operation on the CXL machine, executed by a process}
}
\newglossaryentry{macro step}
{
  name=macro step,
  description={is a sequence of micro steps with at most one instruction
        externally visible to other processes.  It is the unit of interleaving
        in CXL}
}
\newglossaryentry{micro step}
{
  name=micro step,
  description={is the execution of one machine instruction by a process, updating
        its state and the state of the CXL machine}
}
\newglossaryentry{model checking}
{
  name=model checking,
  description={is a formal verification method that explores all possible executions
    of a program, which must have a finite number of states}
}
\newglossaryentry{process}
{
  name=process,
  description={is a method in execution.  We do not make the distinction
      between processes and threads.  A process has a program counter, a register,
      and a stack}
}
\newglossaryentry{process variable}
{
  name=process variable,
  description={is a variable that is private to a single process and stored in
        its register}
}
\newglossaryentry{property}
{
  name=property,
  description={describes a set of execution traces or histories that are allowed by
    a program.  Safety properties are properties in which ``no bad things happen,''
    such as violating mutual exclusion in a critical section.  Liveness properties are
    properties where ``something good eventually happens,'' like processes being
    able to enter the critical section if they want to}
}
\newglossaryentry{race condition}
{
  name=race condition,
  description={describes when multiple processes access shared state concurrently,
    leading to undesirable outcomes}
}
\newglossaryentry{sequential execution}
{
  name=sequential execution,
  description={is when there is just one process executing, as opposed to concurrent
    execution}
}
\newglossaryentry{shared variable}
{
  name=shared variable,
  description={is a variable that is stored in the memory of the CXL machine and
      shared between multiple processes, as opposed to a process variable}
}
\newglossaryentry{stack machine}
{
  name=stack machine,
  description={is a model of computing where the state of a process is kept on
    a stack.  CXL uses a combination of a stack machine and a register-based machine}
}
\newglossaryentry{thread safety}
{
  name=thread safety,
  description={is when the implementation of a data structure allows concurrent access
    with well-defined semantics}
}

@@

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CXL code.
Here \texttt{@cs} is a \emph{label},
\index{label}
identifying a location in the CXL machine code.  The first thing we need to
ensure is that there can never be two processes at the critical section.
This property is called \texttt{mutual exclusion}.
\index{mutual exclusion}
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.

@@

You may already have heard of the concept of a \emph{lock}
\index{lock}
and have realized that
it could be used to implement a critical section.
The idea is that the lock is like a baton that at most one process can own
(or hold) at a time.
A process that wants to enter the critical section at a time must obtain the
lock first and release it upon exiting the critical section.
Using a lock is a good thought, but how does one implement one?
Figure~\ref{fig:uplock} presents a mutual exclusion attempt based on a
na\"{\i}ve (and, as it turns out, broken) implementation of a lock.
Initially the lock is not owned, indicated by lock being \texttt{False}.
To enter the critical section, a process waits until \texttt{lock} is \texttt{False}
and then sets it to \texttt{True} to indicate that the lock has been taken.
The process then executes the critical section.  Finally the process
releases the lock by setting it back to \texttt{False}.

@@

Why does it work?  We will focus here on how one might go about proving
mutual exclusion for an algorithm such as Peterson's.
For that, we have to understand a little bit more about how the CXL
machine works.
In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
\index{state}
at any point of time the CXL machine is in a specific state.
A state is comprised of the values of the shared variables as well as
the values of the process variables
\index{process variable}
\glsadd{process variable}
for each process, including its
program counter and the contents of its stack.
Everytime a process executes a CXL machine instruction, the
state changes (if only because the program counter of the process
changes).  We call that a \emph{step}.
\index{step}
Steps in CXL are atomic.

The CXL machine starts in an initial state in which there is only
one process and its program counter is~0.  A \emph{trace}
\index{trace}
is a sequence of steps starting from the initial state.
When making a step, there are two sources of non-determinism
\index{non-determinism}
in CXL.
One is when
there is more than one process that can make a step.  The other is
when a process executes a \texttt{choose} operation and there is
more than one choice.
Because there is non-determinism, there are multiple possible traces.
We call a state \emph{reachable}
\index{reachable state}
if it is either the initial state
or it can be reached from the initial state through a trace.
A state is final
when there are no processes left to make state changes.

@@

\chapter{Spinlock}
\index{spinlock}

@@

Figure~\ref{fig:uplock} showed a faulty attempt at solving mutual
exclusion using a lock.  The problem with the implementation of the
lock is that checking the lock and setting it if it is available is
not \emph{atomic}.  Thus multiple processes contending for the lock
can all ``grab the lock'' at the same time.  While Peterson's
algorithm gets around the problem, it is not efficient, especially
if generalized to multiple processes.  Instead, multi-core processors provide
so-called \emph{interlock instructions}:
\index{interlock instruction}
special machine instructions
that can read memory and then write it in an indivisible step.

@@

We call a process \emph{blocked}
\index{blocked process}
in a state if the process is in
an \emph{infinite loop} in which the process does not update the
shared memory.
A process trying to
acquire a test-and-set spinlock held by another process is a good example
of a process being blocked: the process only reads the shared memory
but does not change it.
The only way to break out of the infinite loop is if another process
changes the shared memory.
The \texttt{while} loop in Peterson's algorithm, in case its condition
does not hold, is another example of a process being blocked.

@@

\chapter{Reader/Writer Locks using Busy Waiting}
\index{reader/writer lock}

@@

While correct, it is not considered a good solution.
The solution is an example of what is called \emph{busy-waiting}
\index{busy waiting}
(aka spin-waiting):
\index{spin waiting}
processes spin in a loop until some desirable application-level condition is met.
The astute reader might wonder if obtaining the mutex itself is an
example of busy-waiting.  After all, the CXL \texttt{synch} implementation
of \texttt{lock()} spins in a loop until the mutex is available
(see Figure~\ref{fig:spinlocks}).

@@

\chapter{Deadlock}
\index{deadlock}

@@

\chapter{Semaphores}
\index{semaphore}

@@

\chapter{Bounded Buffer}
\index{bounded buffer}
\index{producer/consumer problem}

@@

Reader/writer locks and bounded buffers are both examples of what are
sometimes called \emph{conditional critical sections}
\index{conditional critical section}
(CCSs)~\cite{Hoare73}.
All critical sections are in fact conditional.  Even the basic one
has the condition that a process can only enter if no other process
is in the critical section.  Reader/writer locks refined that, loosening
the restriction for improved performance.
In the case of bounded buffers, there were two CCSs.  First, a CCS
to add an entry to the bounded buffer that can only be entered by one
process and only if there is room in the bounded buffer (\texttt{produce}).
Second, a CCS that can only be entered by one process and
only if there is an item in the bounded buffer (\texttt{consume}).

@@

\chapter{Starvation}
\label{ch:starvation}
\index{starvation}

@@

In this chapter, we will present a version of a reader/writer lock
implementation that solves this type of starvation, but won't eliminate
the type of starvation.
In particular, the probability that either a reader or writer gets starved
will be exponentially diminishing.
We shall call such a solution \emph{fair}
\index{fairness}
(although it does not quite
match the usual formal nor vernacular concepts of fairness).

@@

Tony Hoare, who came up with the concept of split binary semaphores, devised
an abstraction of the concept in a programming language paradigm called
\emph{monitors}~\cite{Hoare74}.
\index{monitor}
(A similar construct was independently invented by Per Brinch Hansen~\cite{BH73}.)
A monitor is a special version of an object-oriented \emph{class}, comprising
a set of variables and methods that operate on those variables.
There is a split binary semaphore associated with each such class.
The \texttt{mutex} is hidden: it is automatically acquired when invoking a
method and released upon exit.
The other semaphores are called \emph{condition variables}.
\index{condition variable}
There are two operations on condition variables: \texttt{wait}
\index{wait}
and
\texttt{signal}.
\index{signal}
If \texttt{c} is a pointer to a condition variable, then the semantics of the
operations, in CXL, are as follows:

@@

\chapter{Actors and Synchronized Queues}
\index{actor}

@@

\chapter{Non-Blocking Synchronization}
\index{non-blocking synchronization}

@@

\chapter{Barrier Synchronization}
\index{barrier synchronization}

@@

A \emph{micro step}
\index{micro step}
\glsadd{micro step}
is the execution of a single CXL machine instruction
by a context.
Each micro step generates a new state.
When there are multiple contexts, the CXL machine can interleave them.
However, trying to interleave every microstep would be needlessly expensive,
as many micro steps involve changes to a context that are invisible to
other contexts.

A \emph{macro step}
\index{macro step}
\glsadd{macro step}
can involve multiple micro steps.  The following
instructions start a new macro step: \texttt{Load}, \texttt{Store},
\texttt{AtomicInc}, and \texttt{Continue}.  The CXL machine
interleaves macro steps, not micro steps.  Like micro steps, each
macro step involves a single context.  Unlike a micro step, a macro
step can leave the state unchanged.

>>>>

\newglossaryentry{actor model}
{
  name=actor model,
  description={is a concurrency model where there are no shared variables, only
      processes with private variables and exchanging messages}
}
\newglossaryentry{atomicity}
{
  name=atomicity,
  description={describes that a certain machine instruction or sequence of machine
    instructions by a process is executed indivisibly and cannot be interleaved with 
    machine instructions of another process}
}
\newglossaryentry{barrier synchronization}
{
  name=barrier synchronization,
  description={is when a set of processes execute in rounds, waiting for one another
    to complete each round}
}
\newglossaryentry{blocked process}
{
  name=blocked process,
  description={is a process cannot make progress without the help of another process.
      For example, a process that is waiting for a spinlock to become available}
}
\newglossaryentry{busy waiting}
{
  name=busy waiting,
  description={(aka spin-waiting) is when a process waits in a loop for some
      application-defined condition while changing the state, for example by
      continually acquiring and releasing a lock}
}
\newglossaryentry{concurrent execution}
{
  name=concurrent execution,
  description={(aka parallel execution) is when there are multiple processes executing and
      their machine instructions are interleaved in an unpredictable manner}
}
\newglossaryentry{condition variable}
{
  name=condition variable,
  description={a variable that keeps track of which processes are waiting for a
    specific application-level condition.  The variable can be waited on as well
    as signaled or notified}
}
\newglossaryentry{context}
{
  name=context,
  description={(aka continuation) describes the state of a running process,
    including its program counter, the values of its variables (stored in its
    register), and the contents of its stack}
}
\newglossaryentry{conditional critical section}
{
  name=conditional critical section,
  description={is a critical section with, besides mutual exclusion, additional conditions on
      when a process is allowed to enter the critical section}
}
\newglossaryentry{critical section}
{
  name=critical section,
  description={(aka critical region) is a set of instructions that only one process
    is allowed to execute at a time.  The instructions are, however, not executed
    atomically, as other processes can continue to execute and access shared
    variables}
}
\newglossaryentry{deadlock}
{
  name=deadlock,
  description={is when there are two or more processes waiting indefinitely for one
    another to release a resource}
}
\newglossaryentry{determinism}
{
  name=determinism,
  description={is when the outcome of an execution is uniquely determined by the
    initial state}
}
\newglossaryentry{fairness}
{
  name=fairness,
  description={is when each process eventually can access each resource it needs to access
    with high probability}
}
\newglossaryentry{interlock instruction}
{
  name=interlock instruction,
  description={a machine instruction that involves multiple memory load and/or store
    operations, executed atomically}
}
\newglossaryentry{invariant}
{
  name=invariant,
  description={is a binary predicate over states that must hold for every
  reachable state of a process}
}
\newglossaryentry{lock}
{
  name=lock,
  description={an object that can be owned by at most one process at a time.  Useful
    for implementing mutual exclusion}
}
\newglossaryentry{machine instruction}
{
  name=machine instruction,
  description={is an atomic operation on the CXL machine, executed by a process}
}
\newglossaryentry{model checking}
{
  name=model checking,
  description={is a formal verification method that explores all possible executions
    of a program, which must have a finite number of states}
}
\newglossaryentry{monitor}
{
  name=monitor,
  description={is a programming language paradigm that supports mutual exclusion
    as well as waiting for resources to become available}
}
\newglossaryentry{mutual exclusion}
{
  name=mutual exclusion,
  description={is a property that two processes never enter the same critical section}
}
\newglossaryentry{non-blocking synchronization}
{
  name=non-blocking synchronization,
  description={(aka wait-free synchronization) is when access to a shared resource can be
      guaranteed in a bounded number of steps even if other processes are not making progress}
}
\newglossaryentry{process}
{
  name=process,
  description={is a method in execution.  We do not make the distinction
      between processes and threads.  A process has a program counter, a register,
      and a stack}
}
\newglossaryentry{process variable}
{
  name=process variable,
  description={is a variable that is private to a single process and stored in
        its register}
}
\newglossaryentry{producer/consumer problem}
{
  name=producer/consumer problem,
  description={is a synchronization problem whereby one or more producing processes
    submit items and one or more consuming process want to receive them.  No item
    can get lost or forged or be delivered to more than one consumer, and producers
    and consumers should block if resources are exhausted}
}
\newglossaryentry{property}
{
  name=property,
  description={describes a set of execution traces or histories that are allowed by
    a program.  Safety properties are properties in which ``no bad things happen,''
    such as violating mutual exclusion in a critical section.  Liveness properties are
    properties where ``something good eventually happens,'' like processes being
    able to enter the critical section if they want to}
}
\newglossaryentry{race condition}
{
  name=race condition,
  description={describes when multiple processes access shared state concurrently,
    leading to undesirable outcomes}
}
\newglossaryentry{reader/writer lock}
{
  name=reader/writer lock,
  description={is a lock on a resource that can be held by multiple processes if they all 
    only read the resource}
}
\newglossaryentry{semaphore}
{
  name=semaphore,
  description={is a counter that can be atomically incremented and decremented,
    but blocks the process until the counter is larger than zero first}
}
\newglossaryentry{sequential execution}
{
  name=sequential execution,
  description={is when there is just one process executing, as opposed to concurrent
    execution}
}
\newglossaryentry{shared variable}
{
  name=shared variable,
  description={is a variable that is stored in the memory of the CXL machine and
      shared between multiple processes, as opposed to a process variable}
}
\newglossaryentry{spinlock}
{
  name=spinlock,
  description={is an implementation of a lock whereby a process loops until the
    lock is available, at which point the process atomically obtains the lock}
}
\newglossaryentry{stack machine}
{
  name=stack machine,
  description={is a model of computing where the state of a process is kept on
    a stack.  CXL uses a combination of a stack machine and a register-based machine}
}
\newglossaryentry{starvation}
{
  name=starvation,
  description={is when a process cannot make progress because it is continuously
    losing a competition with other processes to get access to a resource}
}
\newglossaryentry{state}
{
  name=state,
  description={an assignment of values to variables.  In a CXL machine, this includes
      the contents of its shared memory and the set of contexts}
}
\newglossaryentry{step}
{
  name=step,
  description={is the execution of a machine instruction by a process, updating
        its state.  CXL distinguishes micro steps (machine instructions) and macro
        step (a sequence of instructions with at most one effect visible by
        other processes)}
}
\newglossaryentry{thread safety}
{
  name=thread safety,
  description={is when the implementation of a data structure allows concurrent access
    with well-defined semantics}
}
\newglossaryentry{trace}
{
  name=trace,
  description={is a sequence of steps, starting from an initial state.  An infinite
    trace is also called a \emph{behavior}}
}

@@

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CXL code.
Here \texttt{@cs} is a \emph{label},
\index{label}
identifying a location in the CXL machine code.  The first thing we need to
ensure is that there can never be two processes at the critical section.
This property is called \texttt{mutual exclusion}.
\index{mutual exclusion}
\glsadd{mutual exclusion}
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.

@@

You may already have heard of the concept of a \emph{lock}
\index{lock}
\glsadd{lock}
and have realized that
it could be used to implement a critical section.
The idea is that the lock is like a baton that at most one process can own
(or hold) at a time.
A process that wants to enter the critical section at a time must obtain the
lock first and release it upon exiting the critical section.
Using a lock is a good thought, but how does one implement one?
Figure~\ref{fig:uplock} presents a mutual exclusion attempt based on a
na\"{\i}ve (and, as it turns out, broken) implementation of a lock.
Initially the lock is not owned, indicated by lock being \texttt{False}.
To enter the critical section, a process waits until \texttt{lock} is \texttt{False}
and then sets it to \texttt{True} to indicate that the lock has been taken.
The process then executes the critical section.  Finally the process
releases the lock by setting it back to \texttt{False}.

@@

Why does it work?  We will focus here on how one might go about proving
mutual exclusion for an algorithm such as Peterson's.
For that, we have to understand a little bit more about how the CXL
machine works.
In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
\index{state}
\glsadd{state}
at any point of time the CXL machine is in a specific state.
A state is comprised of the values of the shared variables as well as
the values of the process variables
\index{process variable}
\glsadd{process variable}
for each process, including its
program counter and the contents of its stack.
Everytime a process executes a CXL machine instruction, the
state changes (if only because the program counter of the process
changes).  We call that a \emph{step}.
\index{step}
\glsadd{step}
Steps in CXL are atomic.

The CXL machine starts in an initial state in which there is only
one process and its program counter is~0.  A \emph{trace}
\index{trace}
\glsadd{trace}
is a sequence of steps starting from the initial state.
When making a step, there are two sources of non-determinism
\index{non-determinism}
in CXL.
One is when
there is more than one process that can make a step.  The other is
when a process executes a \texttt{choose} operation and there is
more than one choice.
Because there is non-determinism, there are multiple possible traces.
We call a state \emph{reachable}
\index{reachable state}
if it is either the initial state
or it can be reached from the initial state through a trace.
A state is final
when there are no processes left to make state changes.

@@

\chapter{Spinlock}
\index{spinlock}
\glsadd{spinlock}

@@

Figure~\ref{fig:uplock} showed a faulty attempt at solving mutual
exclusion using a lock.  The problem with the implementation of the
lock is that checking the lock and setting it if it is available is
not \emph{atomic}.  Thus multiple processes contending for the lock
can all ``grab the lock'' at the same time.  While Peterson's
algorithm gets around the problem, it is not efficient, especially
if generalized to multiple processes.  Instead, multi-core processors provide
so-called \emph{interlock instructions}:
\index{interlock instruction}
\glsadd{interlock instruction}
special machine instructions
that can read memory and then write it in an indivisible step.

@@

We call a process \emph{blocked}
\index{blocked process}
\glsadd{blocked process}
in a state if the process is in
an \emph{infinite loop} in which the process does not update the
shared memory.
A process trying to
acquire a test-and-set spinlock held by another process is a good example
of a process being blocked: the process only reads the shared memory
but does not change it.
The only way to break out of the infinite loop is if another process
changes the shared memory.
The \texttt{while} loop in Peterson's algorithm, in case its condition
does not hold, is another example of a process being blocked.

@@

\chapter{Reader/Writer Locks using Busy Waiting}
\index{reader/writer lock}
\glsadd{reader/writer lock}

@@

While correct, it is not considered a good solution.
The solution is an example of what is called \emph{busy-waiting}
\index{busy waiting}
\glsadd{busy waiting}
(aka spin-waiting):
\index{spin waiting}
processes spin in a loop until some desirable application-level condition is met.
The astute reader might wonder if obtaining the mutex itself is an
example of busy-waiting.  After all, the CXL \texttt{synch} implementation
of \texttt{lock()} spins in a loop until the mutex is available
(see Figure~\ref{fig:spinlocks}).

@@

\chapter{Deadlock}
\index{deadlock}
\glsadd{deadlock}

@@

\chapter{Semaphores}
\index{semaphore}
\glsadd{semaphore}

@@

\chapter{Bounded Buffer}
\index{bounded buffer}
\index{producer/consumer problem}
\glsadd{producer/consumer problem}

@@

Reader/writer locks and bounded buffers are both examples of what are
sometimes called \emph{conditional critical sections}
\index{conditional critical section}
(CCSs)~\cite{Hoare73}.
\glsadd{conditional critical section}
All critical sections are, in a way, conditional.  Even the basic critical section
has the condition that a process can only enter if no other process
is in the critical section.  Reader/writer locks refined that, loosening
the restriction for improved performance.
In the case of bounded buffers, there were two CCSs.  First, a CCS
to add an entry to the bounded buffer that can only be entered by one
process and only if there is room in the bounded buffer (\texttt{produce}).
Second, a CCS that can only be entered by one process and
only if there is an item in the bounded buffer (\texttt{consume}).

@@

\chapter{Starvation}
\label{ch:starvation}
\index{starvation}
\glsadd{starvation}

@@

In this chapter, we will present a version of a reader/writer lock
implementation that solves this type of starvation.
In particular, the probability that either a reader or writer gets starved
will be exponentially diminishing.
We shall call such a solution \emph{fair}
\index{fairness}
\glsadd{fairness}
(although it does not quite
match the usual formal nor vernacular concepts of fairness).

@@

Tony Hoare, who came up with the concept of split binary semaphores, devised
an abstraction of the concept in a programming language paradigm called
\emph{monitors}~\cite{Hoare74}.
\index{monitor}
\glsadd{monitor}
(A similar construct was independently invented by Per Brinch Hansen~\cite{BH73}.)
A monitor is a special version of an object-oriented \emph{class}, comprising
a set of variables and methods that operate on those variables.
There is a split binary semaphore associated with each such class.
The \texttt{mutex} is hidden: it is automatically acquired when invoking a
method and released upon exit.
The other semaphores are called \emph{condition variables}.
\index{condition variable}
\glsadd{condition variable}
There are two operations on condition variables: \texttt{wait}
\index{wait}
and
\texttt{signal}.
\index{signal}
If \texttt{c} is a pointer to a condition variable, then the semantics of the
operations, in CXL, are as follows:

@@

\chapter{Actors and Synchronized Queues}
\index{actor model}
\glsadd{actor model}

@@

\chapter{Non-Blocking Synchronization}
\index{non-blocking synchronization}
\glsadd{non-blocking synchronization}

@@

\chapter{Barrier Synchronization}
\index{barrier synchronization}
\glsadd{barrier synchronization}

@@

A \emph{micro step}
is the execution of a single CXL machine instruction
by a context.
Each micro step generates a new state.
When there are multiple contexts, the CXL machine can interleave them.
However, trying to interleave every microstep would be needlessly expensive,
as many micro steps involve changes to a context that are invisible to
other contexts.

A \emph{macro step}
\index{macro step}
can involve multiple micro steps.  The following
instructions start a new macro step: \texttt{Load}, \texttt{Store},
\texttt{AtomicInc}, and \texttt{Continue}.  The CXL machine
interleaves macro steps, not micro steps.  Like micro steps, each
macro step involves a single context.  Unlike a micro step, a macro
step can leave the state unchanged.
